P2P-LLM: Collaborative LLM Inference

### 2025-05-05 docs/paper/main.tex
final

My work explores fully decentralized crowd computing architectures,
where users establish a peer-to-peer network in which each participant can leverage the collective computational resources for individual tasks.
This structure shares conceptual similarities with BitTorrent file-sharing networks.
While current state-of-the-art language models exceed the capacity of individual personal computers,
the aggregation of distributed computational resources through a decentralized network could potentially enable users to host their own LLMs,
thereby democratizing access to advanced artificial intelligence capabilities.
\section{Realted Work}

@@

The idea of pooling compute power from many computers to form a distributed supercomputer is not new. This concept of Crowd Computing
has been used ever since large networks of computer were formed. The predominant use of this method is to facilitate academic research.
Researchers publish problem that that are working on, and anyone can help contribute by volunteering to donate computing power.
Projects that have successfully used crowd computing include SETI@Home \citep{10.1145/581571.581573} and Einstein@Home \citep{Steltner_2021}.
This approach has proven to be successful in the past, but it has become less popular as the computational power of computers increased and access to supercomputers increased.
However, the use of idle compute power still has untapped applications, and the principles from crowd computing can be applied to modern tasks like LLM inference.

@@

LLM inference is a perfect candidate for P2P network because it has technical requirements that prohibit most personal computers from being able to run
large models with billions of parameters, and it can easily be split into small chunks that can be run on several computers at the same time. BitWorker \citep{Durand_Gasparyan_Rouvinez_Aad_Braun_Trinh_2015} is a protocol designed to compute any parallel computing task across a peer-to-peer network.
Although it is possible to apply LLM inference to this protocol, BitWorker is designed for general tasks and would not able to efficiently scale to the demands of many users.
The most notable research being done in this area is Petals \citep{borzunov2023petalscollaborativeinferencefinetuning} which is a BitTorrent like peer-to-peer network for LLM inference and finetuning.
Petals creates a network with client and servers. Clients send requests that will be processed by the servers on the network. The model is split into multiple sequential blocks of layers, and each server
in the network is capable of serving one or multiple of these blocks. In order to run an inference step, a client will find a optimal path of servers
in the network that forms the complete model and minimizes time required. In my work I explore a similar approach with a network of distributed computers that serve a specific part of the whole LLM,
but I expand upon this and add incentive mechanisms that reward nodes for donating their computational power. This incentive mechanism also facilitates optimal allocation of nodes to maximize throughput, by each
node maximizing expected revenue individually.

@@

Fundamentally, a peer-to-peer network is a decentralized system where each devices acts as a client and a sever.
Each devices can send and receive data which eliminates the need for a centralized server. The central benefit of a P2P network
is that resources are pooled and shared between peers, so the combined resources available to an individual peer is greater than that peer's own resources.
A peer-to-peer network can be built to share any type of resource. In the case of BitTorrent, bandwidth and files are shared between peers.
For the case of LLM inference, peers will need to be able to share bandwidth, data, and computational power.

@@

In order to use the network to run the LLM, a peer must first find a sequence of servers in the network that form a complete model.
Because the LLM layers must be computed sequentially, the peer must find a path through the network that will ideally compute
the LLM in the least amount of time possible and at the lowest price. This problem can be solved by viewing the network as a directed graph where nodes are peers serving the LLM lyaers and weights correspond
to the computational cost and latency of nodes. The edges of the graph correspond to the compatibility between the layers that each node in serving.
Specifically, every node is connected to every node that are serving the consecutive layer of the model. A visualization of this directed graph is shown in Figure \ref{fig:layered_graph}.

@@

Due to this property, the P2P network graph is a directed acyclic graph (DAG). This means that there are efficient
algorithms to find the shortest path between two nodes. The first step is to perform a topological sort on the nodes in the network.
This is an ordering of the nodes such that all neighbors of a node come after the node in the ordering.
For any graph $G = (V, E)$ a topological ordering is an order $\prec$ on $V$ such that if $(u \rightarrow v) \in E$ then $u \prec v$.
For example, a topological ordering of the DAG in Figure \ref{fig:layered_graph} is $[0,1,2,3,4,5,6,7,8]$.
The algorithm for topological sort is shown in Appendix \ref{app:topo_sort}. With the topological order, dynamic programming can be used to find the shortest path in the network.
This algorithm will iterate over the ordering, and search all the node's neighbors for the one with the lowest weight.
The full algorithm is shown in Appendix \ref{app:shortest_path}. Most importantly, both the topological sort and path finding algorithm run in linear time
with a computational complexity of $O(|V| + |E|)$.

@@

		\end{axis}
	\end{tikzpicture}
	\caption{Pareto Frontier}
	\label{fig:Pareto-optimal}
\end{figure}
The optimal path is well defined when weights for nodes are clearly defined, but there is not an apparent way to define
weights for nodes in this case. In the P2P-LLM network, there are two competing criterion that define the weight of node: computational cost and price.
The algorithm described above uses a single objective (SO), while the optimal path in the network is a multi-objective (MO) optimization problem.
There are several way to find optimal solutions to MO optimization problems \citep{1599245}. The simplest way is to convert it to a SO optimization problem, and
since the shortest path algorithm is designed for SO objective, it is the most practical solution for this task. In order to reframe the problem as a single objective,
a weighed aggregate of the objectives is used. In the case for the peer-to-peer network, the SO is a weighed sum of computational cost $c_i$ and price $p_i$.
The objective, or weight, of a node, $w_i$ is defined as.

@@

However, this assumes that the parameter $\alpha$ is known a priori. In general, there are many equivalent optimal solutions in a MO optimization problem.
These are known and Pareto-optimal solutions. They are solutions such that there is no way to improve one objective without degrading any other objectives.
A solution is said to be Pareto-optimal if there are no other solutions that dominate it. A solution $x_1$ dominates $x_2$ if
\begin{enumerate}
	\item $\forall i, f_i(x_1) \leq f_i(x_2)$ and
	\item $\exists i, f_i(x_1) < f_i(x_2)$
\end{enumerate}
where $f_i()$ is the utility for the $i$th objective.

@@

Now with the expected revenue function, the optimal price can be found by finding the price that maximizes $R(p_i)$.
Unfortunately, there is not a closed form solution to this problem, but there are methods that can be used.
Iterative pricing mechanism \citep{ecf80b8b-07bb-374f-ab3f-2cc6a349f689} are used to find optimal prices by making iterative changes to the price until it converges at an optimal point.
These methods work well for dynamic environment where the state of the market is continuously changing.
For example, gradient ascent has been used for forex investing with algorithmic trading \citep{su142215328}.
The gradient ascent algorithm is very similar to the gradient descent algorithm \citep{cauchy1847methode} commonly used in machine learning.
The key difference is that it seeks to maximize a function's value rather minimize it.
Gradient ascent can be applied in the P2P-LLM network to maximize the expected revenue function, $R(p_i)$, with respect to price.
It is defined as

@@

where $\mathbf{x}_t$ is the parameter vector at iteration $t$
,$\mathbf{x}_{t+1}$ is the updated parameter vector,
$\eta$ is the learning rate (step size)
, and $\nabla f(\mathbf{x}_t)$ is the gradient of the objective function $f$ at point $\mathbf{x}_t$.

@@

Using this with the gradient ascent algorithm (Equation \ref{eq:grad}), the price update rule is

@@

\section{Cryptoeconomics}
The P2P-LLM network requires a secure, transparent, and decentralized mechanism to facilitate transactions between peers.
The blockchain provides a secure and decentralized way to allow peers to form contracts with eachother
and exchange payment in cryto-currencies for computation.
Cryptoeconomics has proven to be successful in the past with some application being in
resources trading through the blockchain. Specifically, research has been done to enable peer-to-peer surplus energy on the blockchain \citep{WONGTHONGTHAM2021107299}.
For the P2P-LLM network, the blockchain serves three primary functions: (1) recording the state of the network including available nodes and their capabilities, (2) facilitating secure payments between peers, and (3) executing smart contracts that govern computational resource sharing.

@@

For the simulation, I chose to create a small network, with $n = 28$ total nodes, and a simulated LLM with $m = 4$ layers.
The parameters of the nodes in the network were randomly initialized
where $c_i \in (1,50)$, $l_i \in (1,20)$. All prices are intitially set at $1$. The inital state of the network not optimal with the peers being distributed between the $4$ layers in the configuration $[10, 4, 10, 4]$.
For simplicity, all nodes have an $\alpha$ value of $0.5$ and the $\eta$ used in the iterative pricing function is $1$.

@@

When running the simulated environment, the behavior of the peers aligns with what is expected based on the theory described above.
The starting state of the network has uniform pricing across all nodes, and there is a high concentration of peers in the first and third layers of the LLM.
As the network progresses toward a state of equilibirum, the prices are expected to rise to meet the demand of the network, and reach a maximum where the expected revenue for the reective node is highest.
Nodes are also expected to change the layer of the LLM they serve based on the location of other nodes in the network. Ideally, the nodes will be evenly distriubted between layers in the network.
Looking at Table \ref{tab:network-metrics}, the expected actions occurs as the iteration of the simulation increases.
It eventually reached a point of equilibrium where the prices and layers are constant.

>>>>

My work explores fully decentralized crowd computing architectures,
where users establish a peer-to-peer network in which each participant can leverage the collective computational resources for individual tasks.
This structure shares conceptual similarities with BitTorrent file-sharing networks.
While current state-of-the-art language models exceed the capacity of individual personal computers,
the aggregation of distributed computational resources through a decentralized network could potentially enable users to host their own LLMs,
thereby democratizing access to advanced artificial intelligence capabilities.
\section{Related Work}

@@

The idea of pooling compute power from many computers to form a distributed supercomputer is not new. This concept of Crowd Computing
has been used ever since large networks of computer were formed. The predominant use of this method is to facilitate academic research.
Researchers publish problem that that they are working on, and anyone can help contribute by donating their computing power.
Projects that have successfully used crowd computing include SETI@Home \citep{10.1145/581571.581573} and Einstein@Home \citep{Steltner_2021}.
This approach has proven to be successful in the past, but it has become less popular as the computational power of computers and access to supercomputers has increased.
However, the use of idle compute power still has untapped applications, and the principles from crowd computing can be applied to modern tasks like LLM inference.

@@

LLM inference is a perfect candidate for P2P network because it has technical requirements that prohibit most personal computers from being able to run
large models with billions of parameters, and it can easily be split into small chunks that can be run on several computers at the same time. BitWorker \citep{Durand_Gasparyan_Rouvinez_Aad_Braun_Trinh_2015} is a protocol designed to compute any parallel computing task across a peer-to-peer network.
Although it is possible to apply LLM inference to this protocol, BitWorker is designed for general tasks and would not able to efficiently scale to the demands of many users.
The most notable research being done in this area is Petals \citep{borzunov2023petalscollaborativeinferencefinetuning} which is a BitTorrent like peer-to-peer network for LLM inference and finetuning.
Petals creates a network with client and servers. Clients send requests that will be processed by the servers on the network. The model is split into multiple sequential blocks of the LLM's layers, and each server
in the network is capable of serving one or multiple of these blocks. In order to run an inference step, a client will find a optimal path of servers
in the network that forms the complete model and minimizes the time required to compute the LLM. In my work, I explore a similar approach with a network of distributed computers that serve a specific part of the LLM,
but I expand upon this and add incentive mechanisms that reward nodes for donating their computational power. This incentive mechanism also facilitates optimal allocation of nodes to maximize throughput, by each
node maximizing expected revenue individually.

@@

Fundamentally, a peer-to-peer network is a decentralized system where each devices acts as a client and a server.
Each devices can send and receive data which eliminates the need for a centralized server. The central benefit of a P2P network
is that resources are pooled and shared between peers, so the combined resources available to an individual peer is greater than that peer's own resources.
A peer-to-peer network can be built to share any type of resource. In the case of BitTorrent, bandwidth and files are shared between peers.
For the case of LLM inference, peers will need to be able to share bandwidth, data, and computational power.

@@

In order to use the network to run the LLM, a peer must first find a sequence of servers in the network that form a complete model.
Because the LLM layers must be computed sequentially, the peer must find a path through the network that will ideally compute
the LLM in the least amount of time possible and at the lowest price. This problem can be solved by viewing the network as a directed graph where nodes are peers serving the LLM layers and weights correspond
to the computational cost and latency of nodes. The edges of the graph correspond to the compatibility between the layers that each node in serving.
Specifically, each node is connected to every node that is serving the consecutive layer of the model. A visualization of this directed graph is shown in Figure \ref{fig:layered_graph}.

@@

Due to this property, the P2P network graph is a directed acyclic graph (DAG). This means that there are efficient
algorithms to find the shortest path between two nodes. The first step is to perform a topological sort on the nodes in the network.
This is an ordering of the nodes such that all neighbors of a node come after the node in the ordering.
For any graph $G = (V, E)$ a topological ordering is an order $\prec$ on $V$ such that if $(u \rightarrow v) \in E$ then $u \prec v$.
For example, a topological ordering of the DAG in Figure \ref{fig:layered_graph} is $[0,1,2,3,4,5,6,7,8]$.
The complete algorithm for topological sort is shown in Appendix \ref{app:topo_sort}. With the topological order, dynamic programming can be used to find the shortest path in the network.
This algorithm will iterate over the ordering, and search all the node's neighbors for the one with the lowest weight.
The full algorithm is shown in Appendix \ref{app:shortest_path}. Most importantly, both the topological sort and path finding algorithm run in linear time
with a computational complexity of $O(|V| + |E|)$.

@@

		\end{axis}
	\end{tikzpicture}
	\caption{Pareto Frontier}
	\label{fig:Pareto-optimal}
\end{figure}
The optimal path is well defined when weights for nodes are clearly defined, but there is not an apparent way to define
weights for nodes in this case. In the P2P-LLM network, there are two competing criterion that define the weight of node: computational cost and price.
The algorithm described above uses a single objective (SO), while the optimal path in the network is a multi-objective (MO) optimization problem.
There are several way to find optimal solutions to MO optimization problems \citep{1599245}. The simplest way is to convert it to a SO optimization problem, and
since the shortest path algorithm is designed for a single objective, it is the most practical solution for this task. In order to reframe the problem as a single objective,
a weighed aggregate of the multiple objectives is used. In the case of the peer-to-peer network, the SO is a weighed sum of computational cost $c_i$ and price $p_i$.
The objective, or weight, of a node $w_i$ is defined as.

@@

However, this assumes that the parameter $\alpha$ is known a priori. In general, there are many equivalent optimal solutions in a MO optimization problem.
These are known as Pareto-optimal solutions. They are solutions such that there is no way to improve one objective without degrading any other objectives.
A solution is said to be Pareto-optimal if there are no other solutions that dominate it. A solution $x_1$ dominates $x_2$ if
\begin{enumerate}
	\item $\forall i, f_i(x_1) \leq f_i(x_2)$ and
	\item $\exists i, f_i(x_1) < f_i(x_2)$
\end{enumerate}
where $f_i()$ is the utility for the $i$th objective.

@@

Now with the expected revenue function, the optimal price can be found by finding the price that maximizes $R(p_i)$.
Unfortunately, there is not a closed form solution to this problem, but there are methods that can be used.
Iterative pricing mechanism \citep{ecf80b8b-07bb-374f-ab3f-2cc6a349f689} are used to find optimal prices by making iterative changes to the price until it converges at an optimal point.
These methods work well for dynamic environment where the state of the market is continuously changing.
For example, gradient ascent has been used for forex investing with algorithmic trading \citep{su142215328}.
The gradient ascent algorithm is very similar to the gradient descent algorithm \citep{cauchy1847methode} commonly used in machine learning.
The key difference is that it seeks to maximize a function's value rather than minimize it.
Gradient ascent can be applied in the P2P-LLM network to maximize the expected revenue function, $R(p_i)$, with respect to price.
It is defined as

@@

where $\mathbf{x}_t$ is the parameter vector at iteration $t$,
$\mathbf{x}_{t+1}$ is the updated parameter vector,
$\eta$ is the learning rate (step size),
and $\nabla f(\mathbf{x}_t)$ is the gradient of the objective function $f$ at point $\mathbf{x}_t$.

@@

Using this with the gradient ascent algorithm (Equation \ref{eq:grad}), the price update rule becomes

@@

\section{Cryptoeconomics}
The P2P-LLM network requires a secure, transparent, and decentralized mechanism to facilitate transactions between peers.
The blockchain provides a secure and decentralized way to allow peers to form contracts with eachother
and exchange payment in cryto-currencies for computation.
Cryptoeconomics has proven to be successful in the past with some application being in
resources trading through the blockchain. Specifically, research has been done to enable peer-to-peer surplus energy trading on the blockchain \citep{WONGTHONGTHAM2021107299}.
For the P2P-LLM network, the blockchain serves three primary functions: (1) recording the state of the network including available nodes and their capabilities, (2) facilitating secure payments between peers, and (3) executing smart contracts that govern computational resource sharing.

@@

For the simulation, I chose to create a small network, with $n = 28$ total nodes, and a simulated LLM with $m = 4$ layers.
The parameters of the nodes in the network were randomly initialized
where $c_i \in (1,50)$, $l_i \in (1,20)$. All prices are initially set at $1$. The initial state of the network is not optimal with the peers being distributed between the $4$ layers in the configuration $[10, 4, 10, 4]$.
For simplicity, all nodes have an $\alpha$ value in Equation \ref{eq:weight} of $0.5$ and the $\eta$ used in the iterative pricing function (Equation \ref{eq:optim_price}) is $1$.

@@

When running the simulated environment, the behavior of the peers aligns with what is expected based on the theory described above.
The starting state of the network has uniform pricing across all nodes, and there is a high concentration of peers in the first and third layers of the LLM.
As the network progresses toward a state of equilibrium, the prices are expected to rise to meet the demand of the network, and reach a maximum where the expected revenue for the respective node is highest.
Nodes are also expected to change the layer of the LLM they serve based on the location of other nodes in the network. Ideally, the nodes will be evenly distributed between layers in the network.
Looking at Table \ref{tab:network-metrics}, the expected actions occurs as the iteration of the simulation increases.
It eventually reached a point of equilibrium where the prices and layers are constant.
